Diagnoses and cell types

### 2025-08-21 analyses/infercnv-consensus-cell-type/exploratory-notebooks/reference-query-counts.Rmd
fix counting to handle multiplexed samples, and update language throughout to libraries which is actually what we are counting

The goal of this notebook is to look at cell type distributions across samples to determine how many samples could be run through `inferCNV`.
Specifically, we are interested in seeing the relative numbers of cells intended as `inferCNV` reference vs. query.
Cell types were determined to be reference or query based on the sample's diagnosis, as mapped out in `../references/broad-diagnosis-map.tsv`.
We consider a threshold of at least `r params$n_reference_threshold` reference cells needed to run `inferCNV`.

@@

```{r}
all_sample_diagnosis_df <- metadata_files |>
  purrr::map(
    \(file) {
      readr::read_tsv(file) |>
        dplyr::select(contains("scpca"), diagnosis, is_cell_line, is_multiplexed)
  }) |>
  dplyr::bind_rows() |>
  # use inner_join to remove normals
  dplyr::inner_join(diagnosis_map, by = c("diagnosis" = "submitted_diagnosis")) |>
  dplyr::select(-ontology_id, -human_readable_value)

total_samples <- nrow(all_sample_diagnosis_df) # needed for later math

sample_diagnosis_df <- all_sample_diagnosis_df |>
  # don't consider cell line (or multiplexed?)
  # for now at least, we are retaining multiplexed
  dplyr::filter(!is_cell_line)
```

```{r}
consensus_df <- consensus_files |>
  # only keep samples in sample_diagnosis_df
  purrr::keep_at(sample_diagnosis_df$scpca_library_id) |>
  purrr::imap(
    \(file, id) {

      # bump guess_max to kill problems() warnings
      readr::read_tsv(file, guess_max = 15000) |>
       dplyr::select(
         scpca_project_id = project_id,
         scpca_sample_id = sample_id,
         scpca_library_id = library_id,
         barcodes,
         consensus_annotation,
         consensus_ontology
       )
  }) |>
  dplyr::bind_rows()
```

@@

```{r}
query_reference_df <- unique(sample_diagnosis_df$diagnosis_group) |>
  purrr::map(
    \(broad_diagnosis) {

@@

      library_ids <- sample_diagnosis_df |>
        dplyr::filter(diagnosis_group == broad_diagnosis) |>
        dplyr::pull(scpca_library_id)

@@

```{r}
sample_counts_df <- query_reference_df |>
  dplyr::count(scpca_library_id, cell_class) |>
  tidyr::pivot_wider(
    names_from = cell_class, 
    values_from = n, 
    values_fill = 0
  ) |>
  dplyr::inner_join(sample_diagnosis_df, by = "scpca_library_id") |>
  dplyr::select(
    scpca_project_id, 
    scpca_library_id, 
    diagnosis_group, 
    diagnosis, 
    query, 
    reference
  )
```

@@

Using a threshold of needing at least `r params$n_reference_threshold` reference cells, how many samples could be run through `inferCNV`?

```{r}
usable_df <- sample_counts_df |>
  dplyr::group_by(diagnosis_group) |>
  dplyr::summarize(
    n_usable = sum(reference >= params$n_reference_threshold),
    total_group_samples = dplyr::n(), 
    frac_usable = round(n_usable / total_group_samples, 3) 
  ) |>
  dplyr::arrange(desc(total_group_samples)) |>
  dplyr::select(diagnosis_group, total_group_samples, n_usable, frac_usable)

usable_df
```
All of our broad diagnosis groups are represented, which is nice to see.

Which diagnosis groups have _no samples_ with sufficient reference cells?

```{r}
usable_df |>
  dplyr::filter(frac_usable == 0)
```
It's not too surprising to see `Mixed phenotype acute leukemia` here since most immune cell types can't be included in the reference, and the other groups comprise very small numbers of samples.

In total, this fraction of ScPCA samples have sufficient reference cells to run through `inferCNV`:

```{r}
sum(sample_counts_df$reference >= params$n_reference_threshold) / total_samples
```

Of the samples with sufficient reference cells, do any have _fewer_ query than reference cells?

```{r}
sample_counts_df |>
  # consider usable only
  dplyr::filter(reference >= params$n_reference_threshold) |>
  dplyr::filter(query < params$n_reference_threshold)
```

@@

This sample indeed is dominated by immune cell types with very low tumor purity, but it still seems to have enough query cells to not be a problem and should be ok to run.

@@

In this section, we plot the relative contributions of each cell type group in the samples we can run. 
We show the relative proportions of consensus cell type groups in the _reference_, and we show all other query cells grouped together in gray.
The x-axis is library ids, and the y-axis is the proportion of each cell type group.

```{r}
usable_libraries <- sample_counts_df |>
  dplyr::filter(reference >= params$n_reference_threshold) |>
  dplyr::pull(scpca_library_id)

@@

Currently, _before_ incorporating `SCimilarity` into our consensus cell types, we are able to run about 44% of samples through `inferCNV`. 
Most broad diagnosis groups are represented, except `Mixed phenotype acute leukemia` and a handful of other solid tumors.
All eligible projects except `SCPCP000011` (Retinoblastoma samples) are represented as well.

>>>>

The goal of this notebook is to look at cell type distributions across libraries to determine how many could be run through `inferCNV`.
Specifically, we are interested in seeing the relative numbers of cells intended as `inferCNV` reference vs. query.
Cell types were determined to be reference or query based on the sample's diagnosis, as mapped out in `../references/broad-diagnosis-map.tsv`.
We consider a threshold of at least `r params$n_reference_threshold` reference cells needed to run `inferCNV`.

@@

```{r}
 all_library_diagnosis_df <- metadata_files |>
  purrr::map(
    \(file) {
      df <- readr::read_tsv(file) |>
        dplyr::select(scpca_project_id, scpca_library_id, diagnosis, is_cell_line, is_multiplexed) |>
        # have to join here because of multiplexed; inner_join also removes normals
        dplyr::inner_join(diagnosis_map, by = c("diagnosis" = "submitted_diagnosis")) |>
        dplyr::select(-diagnosis, -ontology_id, -human_readable_value) |>
        dplyr::distinct()
  }) |>
  dplyr::bind_rows()

total_libraries <- nrow(all_library_diagnosis_df) # needed for later math

library_diagnosis_df <- all_library_diagnosis_df |>
  # don't consider cell line (or multiplexed?)
  # for now at least, we are retaining multiplexed
  dplyr::filter(!is_cell_line)
```

```{r}
consensus_df <- consensus_files |>
  # only keep libraries in scpca_project_id_diagnosis_df
  purrr::keep_at(library_diagnosis_df$scpca_library_id) |>
  purrr::imap(
    \(file, id) {

      # bump guess_max to kill problems() warnings
      readr::read_tsv(file, guess_max = 15000) |>
       dplyr::select(
         scpca_project_id = project_id,
         scpca_library_id = library_id,
         barcodes,
         consensus_annotation,
         consensus_ontology
       )
  }) |>
  dplyr::bind_rows()
```

@@

```{r}
query_reference_df <- unique(library_diagnosis_df$diagnosis_group) |>
  purrr::map(
    \(broad_diagnosis) {

@@

      library_ids <- library_diagnosis_df |>
        dplyr::filter(diagnosis_group == broad_diagnosis) |>
        dplyr::pull(scpca_library_id)

@@

```{r}
library_counts_df <- query_reference_df |>
  dplyr::count(scpca_library_id, cell_class) |>
  tidyr::pivot_wider(
    names_from = cell_class, 
    values_from = n, 
    values_fill = 0
  ) |>
  dplyr::inner_join(library_diagnosis_df, by = "scpca_library_id") |>
  dplyr::select(
    scpca_project_id, 
    scpca_library_id, 
    diagnosis_group, 
    query, 
    reference
  )
```

@@

Using a threshold of needing at least `r params$n_reference_threshold` reference cells, how many libraries could be run through `inferCNV`?

```{r}
usable_df <- library_counts_df |>
  dplyr::group_by(diagnosis_group) |>
  dplyr::summarize(
    n_usable = sum(reference >= params$n_reference_threshold),
    total_group_libs = dplyr::n(), 
    frac_usable = round(n_usable / total_group_libs, 3) 
  ) |>
  dplyr::arrange(desc(total_group_libs)) |>
  dplyr::select(diagnosis_group, total_group_libs, n_usable, frac_usable)

usable_df
```
Except for a handful of "other solid tumors," our broad diagnosis groups are well-represented, which is nice to see.

Which diagnosis groups have _no libraries_ with sufficient reference cells?

```{r}
usable_df |>
  dplyr::filter(frac_usable == 0)
```
It's not too surprising to see `Mixed phenotype acute leukemia` here since most immune cell types can't be included in the reference, and the other groups comprise very small numbers of libraries.

In total, this fraction of ScPCA libraries have sufficient reference cells to run through `inferCNV`:

```{r}
sum(library_counts_df$reference >= params$n_reference_threshold) / total_libraries
```

Of the libraries with sufficient reference cells, do any have _fewer_ query than reference cells?

```{r}
library_counts_df |>
  # consider usable only
  dplyr::filter(reference >= params$n_reference_threshold) |>
  dplyr::filter(query < params$n_reference_threshold)
```

@@

This library indeed is dominated by immune cell types with very low tumor purity, but it still seems to have enough query cells to not be a problem and should be ok to run.

@@

In this section, we plot the relative contributions of each cell type group in the libraries we can run. 
We show the relative proportions of consensus cell type groups in the _reference_, and we show all other query cells grouped together in gray.
The x-axis is library ids, and the y-axis is the proportion of each cell type group.

```{r}
usable_libraries <- library_counts_df |>
  dplyr::filter(reference >= params$n_reference_threshold) |>
  dplyr::pull(scpca_library_id)

@@

Currently, _before_ incorporating `SCimilarity` into our consensus cell types, we are able to run about 44% of libraries through `inferCNV`. 
Most broad diagnosis groups are represented, except `Mixed phenotype acute leukemia` and a handful of other solid tumors.
All eligible projects except `SCPCP000011` (Retinoblastoma samples) are represented as well.
